Supplementary Information

### 2021-03-18 doc/manuscript/ForestGEO_treerings.Rmd
remove unnecessary latex
#121

RW_interactions_all_legend = "Figure S56. All signficant climate - DBH interactions with $RW$ as the response metric."
BAI_interactions_all_legend ="Figure S57. All signficant climate - DBH interactions with $RW$ as the response metric."

decadal_plot_BCNM_legend = "Figure S58. Relationships between DBH and $BAI$ by decade for BCNM."
decadal_plot_HKK_legend = "Figure S59. Relationships between DBH and $BAI$ by decade for HKK."
decadal_plot_SCBI_legend = "Figure S60. Relationships between DBH and $BAI$ by decade for SCBI."
decadal_plot_LDW_legend = "Figure S61. Relationships between DBH and $BAI$ by decade for LDW."
decadal_plot_HF_legend = "Figure S62. Relationships between DBH and $BAI$ by decade for HF."
decadal_plot_ZOF_legend = "Figure S63. Relationships between DBH and $BAI$ by decade for ZOF."
decadal_plot_NIO_legend = "Figure S64. Relationships between DBH and $BAI$ by decade for NIO."
decadal_plot_LT_legend = "Figure S65. Relationships between DBH and $BAI$ by decade for LT."
decadal_plot_CB_legend = "Figure S66. Relationships between DBH and $BAI$ by decade for CB."
decadal_plot_SC_legend = "Figure S67. Relationships between DBH and $BAI$ by decade for SC."

@@

*(keep mostly the same):*
Tree diameter at breast height (DBH) scales predictably with numerous variables affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
However, for dendrochronological studies aimed at deciphering climate signals, DBH is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
Moreover, many studies constrain sampling to only larger size classes.
While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of ring width ($RW$) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_agedependent_2007; @rollinson_climate_2021].
However, the removal of DBH signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
To use tree rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].

@@

All tree cores **(n = XXXX)** were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
From among the full set of original $RW$ measurements, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis. 
We also excluded records with insufficient sample size or anomalous growth patterns, including
(1) species with < 7 cores,
(2) cores with < 30 years of record,
(3) contiguous portions of cores containing large outliers ($RW$ > mean plus 5 x SD of $RW$ for the entire core),
and (4) the final 20 years prior to death of trees cored dead. 
The final criterion was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
From analyses including DBH (see below), we further excluded
(1) trees for which we lacked data required to reconstruct DBH,
(2) trees for which there was a significant inconsistency between measured DBH and the sum of $RW$'s across the core (Appendix `r DBH_reconstruction`),
and (3) poorly represented tails of the DBH distribution, starting where reconstructed DBH (see below) was represented by < 3 conspecific trees.
In total, this resulted in inclusion of `r n_cores_all` cores from `r n_trees_all` trees, `r n_cores_dbh` of which (from `r n_trees_dbh` trees) could be included in analyses with DBH (Table `r sampling_details_table`).

@@

Once DBH had been reconstructed, we estimated basal area ($BA=\pi r^2$, where $r$ is radius) and aboveground biomass ($AGB$). 
Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in revision) and *biomass* [@rejou-mechain_biomass_2017], respectively.
We then calculated basal area increment ($BAI= BA_{y+1}-BA_y$, where $y$ is year) and aboveground biomass growth increments ($\Delta AGB=AGB_{y+1}-AGB_y$].

@@

Our analysis consisted of two main steps: (1) identification of the primary climate drivers , and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig. `r schematic`). 
The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable ($RW$, $BAI$, or $\Delta AGB$). 
We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (Fig. `r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.

![**Figure 1 | Schematic illustrating our analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, $RW$) for all cores sampled at the site. In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.](tables_figures/schematic_figure.png)

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals. 
The smoothing parameters were automaticaly selected by the *gam* function by generalized cross-validation (GCV).
We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $RW$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects (noting that these effects should be minimal given that residuals are centered around zero): `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
Here, for each permutation, `climate` specifies one of climate variables within each group (water or temperature), analyzed over one of all possible combinations of consecutive months over a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables, we moved forward with the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.

@@

Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014] and removed any variable with a variance inflation factor > 3 (none required removal).
Within the GLS models, our response variables were $log[\Delta RW]$, $log[BAI]$, or $log[\Delta AGB]$.

@@

As an example, a full model for $log[RW]$ would look like this in *R*:

*lme($log[RW]$ ~ PET + I(PET^2) + $T_{min}$ + I($T_{min}$^2) + DBH + I(DBH^2)",  random = ~1|coreID, correlation = corCAR1(form=~yeaar|coreID), data = x, na.action = "na.fail", method = "ML")*

where $x$ is a complete data set (with no missing value) for one species at one site. The method is set to maximum likelihood (*ML*) during the model selection phase, but to restricted maximizing likelihood (*REML*) once the best model is identified.

@@

To test for yeaar effects, we limited the analysis to species with reasonable coverage of the DBH x `year` matrix.
Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total DBH range for $\ge 2/3$ of the total time range analyzed. 
To avoid severe big-tree selection biases [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm, if present.
Species that failed to meet these criteria (n= 8) were excluded from the analysis of temporal trends.
We then ran models as described above, including a  first-order linear effect of year.
We note that the random effect of tree should in theory avoid analytical biases arising from persistent growth differences among individuals [@brienen_detecting_2012; @brienen_tree_2017]. 
To verify that GLS model trends for yeaar were real, as opposed to an artifact of inherent covariation between DBH and yeaar within each core, we compared GLS results to an analysis of DBH-growth relationships by decade (Figs. `r decadal_analysis_figs_all`). 
For simplicity and consistency with previous literature, we present only the models with $BAI$ as the growth metric, noting that responses were similar across growth metrics.

Within each of three categories of models run (climate only, climate + DBH, climate $\times DBH$, climate + DBH + yeaar), we selected as the top model that with the lowest AIC.

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration (PET). Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R, @zang_dendroclimatic_2013]. Black rectangle represents the period selected by *climwin* as the most influential window. Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta AIC$ for each. The time window with the minimum $\Delta AIC$ (1-3 months prior to August, or May-July; black circles), was identified as the optimal window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model). Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Species names corresponding to the codes are given in Table `r species_table`.](tables_figures/quilt_comparison.png)

@@

```{r , eval=FALSE, echo=FALSE}
## Tallies for the following paragraph (Feb 11)
#sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
#sameV_diffW: BCNMp, BCNMt, HFt, HKKp, LDWt,  SCBIt, SCp,SCt,  CBp, CBt (10)
#diffV_diffW: HFp, HKKt
```
The three metrics of growth ($RW$, $BAI$, and $\Delta AGB$)  exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs. `r all_climwin`).
In 8 of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_SCBI_W`). 
In 10 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_SCBI_T`). 
Finally, in two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF), *climwin* identified different climate variables and different time windows (Figs. `r climwin_HKK_T`, `r climwin_HF_T`). 
Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with $RW$ as the growth metric and for the full set of cores (*i.e.*, including those for which DBH could not be reconstructed).

@@

![**Figure 3 | Species-level responses of $RW$ to climwin-selected variables in precipitation and temperature variable groups.** *Climwin*-selected climate variables are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential. For each species (color-coded as in Fig. 4), relationships are plotted if included in top model. For each relationship shown, other terms in the model are held constant at their median. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)

@@

Interactive effects of climate and DBH were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-variable combinations for which they were tested. 
For precipitation variables, interactions were significant for `r n_DBHxP_RW_sig` of the `r n_DBHxP_RW_analyzed` (`r round(n_DBHxP_RW_sig/ n_DBHxP_RW_analyzed*100)`%) interactions with $RW$ as the growth metric (Fig. `r RW_interactions_all`) and for `r n_DBHxP_BAI_sig` of the `r n_DBHxP_BAI_analyzed` (`r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`%) with $BAI$ as the growth metric (Fig. `r BAI_interactions_all`).
The majority of these interactions were positive (`r round(n_DBHxP_RW_sig_pos/n_DBHxP_RW_sig*100)`% for $RW$; `r round(n_DBHxP_BAI_sig_pos/n_DBHxP_BAI_sig*100)`% for $BAI$), indicating that larger trees generally respond more positively to precipitation or its frequency (Fig.  `r interactions_fig`).
Among the exceptions to this pattern (n=`r n_DBHxP_RW_sig-n_DBHxP_RW_sig_pos` for $RW$, `r n_DBHxP_BAI_sig-n_DBHxP_BAI_sig_pos` for $BAI$), only a minority (n=1 for $RW$, 4 for $BAI$) occurred in species responding positively to precipitation in the current growing season.

Temperature variable $\times DBH$ interactions were significant for `r round(n_DBHxT_RW_sig/ n_DBHxT_RW_analyzed*100)`% of cases with $RW$ as the growth metric (Fig. `r RW_interactions_all`) and for `r round(n_DBHxT_BAI_sig/ n_DBHxT_BAI_analyzed*100)`% with $BAI$ as the growth metric (Fig. `r BAI_interactions_all`).
Directions of these interactions were mixed, with `r n_DBHxT_RW_sig_neg` of `r n_DBHxT_RW_sig` significant interactions negative with $RW$ as the growth metric and `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative when $BAI$ was the growth metric.
For both $RW$ and $BAI$, the majority of significant negative interactions (*i.e.*, more negative/ less positive response of larger trees to higher temperatures) occurred in cases where the main effect temperature was negative (e.g., HKK, LT, CB; Fig.  `r interactions_fig`), whereas positive interactions were more common when the main effect of temperature was positive (e.g., HF, ZOF).

@@

When a precipitation variable, a temperature variable, and DBH were all included as candidate variables in the GLS models, typically all three were included in the top model, regardless of the growth metric used.
Climate responses were generally similar to those described above for models without a DBH term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCNM; Figs. `r clim_sens`, `r all_GLS`).
In general, DBH explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r all_GLS`).
The relative importance of DBH tended to be least for $RW$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. `r GLS_SCBI`).
However, there were exceptions, particularly when $RW$ decreased steeply with DBH (e.g., LT; Fig. `r GLS_LT`).

@@

All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with DBH for most species at all sites (Fig. `r size`).
For $RW$, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear DBH term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative). 
There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
On one end of the spectrum, some species exhibited maximum $RW$ at low DBH, followed by fairly rapid declines in $RW$ with increasing DBH. 
Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) and/or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig. `r size`).
At the other end of the spectrum, some species had low $RW$ at small DBH, increased to peak $RW$ at intermediate DBH, and subsequently declined.
These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp. at SCBI and Žofín, *Acer saccharum* at LDW, *Picea* spp. at Žofín and CB; Table `r species_table`).

>>>>

RW_interactions_all_legend = "Figure S56. All signficant climate - DBH interactions with RW as the response metric."
BAI_interactions_all_legend ="Figure S57. All signficant climate - DBH interactions with RW as the response metric."

decadal_plot_BCNM_legend = "Figure S58. Relationships between DBH and BAI by decade for BCNM."
decadal_plot_HKK_legend = "Figure S59. Relationships between DBH and BAI by decade for HKK."
decadal_plot_SCBI_legend = "Figure S60. Relationships between DBH and BAI by decade for SCBI."
decadal_plot_LDW_legend = "Figure S61. Relationships between DBH and BAI by decade for LDW."
decadal_plot_HF_legend = "Figure S62. Relationships between DBH and BAI by decade for HF."
decadal_plot_ZOF_legend = "Figure S63. Relationships between DBH and BAI by decade for ZOF."
decadal_plot_NIO_legend = "Figure S64. Relationships between DBH and BAI by decade for NIO."
decadal_plot_LT_legend = "Figure S65. Relationships between DBH and BAI by decade for LT."
decadal_plot_CB_legend = "Figure S66. Relationships between DBH and BAI by decade for CB."
decadal_plot_SC_legend = "Figure S67. Relationships between DBH and BAI by decade for SC."

@@

*(keep mostly the same):*
Tree diameter at breast height (DBH) scales predictably with numerous variables affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
However, for dendrochronological studies aimed at deciphering climate signals, DBH is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
Moreover, many studies constrain sampling to only larger size classes.
While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of ring width (RW) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_agedependent_2007; @rollinson_climate_2021].
However, the removal of DBH signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
To use tree rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].

@@

All tree cores **(n = XXXX)** were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
From among the full set of original RW measurements, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis. 
We also excluded records with insufficient sample size or anomalous growth patterns, including
(1) species with < 7 cores,
(2) cores with < 30 years of record,
(3) contiguous portions of cores containing large outliers (RW > mean plus 5 x SD of RW for the entire core),
and (4) the final 20 years prior to death of trees cored dead. 
The final criterion was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
From analyses including DBH (see below), we further excluded
(1) trees for which we lacked data required to reconstruct DBH,
(2) trees for which there was a significant inconsistency between measured DBH and the sum of RW's across the core (Appendix `r DBH_reconstruction`),
and (3) poorly represented tails of the DBH distribution, starting where reconstructed DBH (see below) was represented by < 3 conspecific trees.
In total, this resulted in inclusion of `r n_cores_all` cores from `r n_trees_all` trees, `r n_cores_dbh` of which (from `r n_trees_dbh` trees) could be included in analyses with DBH (Table `r sampling_details_table`).

@@

Once DBH had been reconstructed, we estimated basal area ($BA=\pi r^2$, where $r$ is radius) and aboveground biomass (AGB). 
Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in revision) and *biomass* [@rejou-mechain_biomass_2017], respectively.
We then calculated basal area increment ($BAI= BA_{y+1}-BA_y$, where $y$ is year) and aboveground biomass growth increments ($\Delta AGB=AGB_{y+1}-AGB_y$].

@@

Our analysis consisted of two main steps: (1) identification of the primary climate drivers , and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig. `r schematic`). 
The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable (RW, BAI, or $\Delta$AGB). 
We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (Fig. `r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.

![**Figure 1 | Schematic illustrating our analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site. In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.](tables_figures/schematic_figure.png)

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, thereby producing residuals. 
The smoothing parameters were automaticaly selected by the *gam* function by generalized cross-validation (GCV).
We then used *climwin* to identify the climate variables most strongly correlated to the residuals of RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects (noting that these effects should be minimal given that residuals are centered around zero): `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
Here, for each permutation, `climate` specifies one of climate variables within each group (water or temperature), analyzed over one of all possible combinations of consecutive months over a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables, we moved forward with the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.

@@

Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014] and removed any variable with a variance inflation factor > 3 (none required removal).
Within the GLS models, our response variables were log[RW], log[BAI], or log[$\Delta$ AGB].

@@

As an example, a full model for log[RW] would look like this in *R*:

*lme(log[RW] ~ PET + I(PET^2) + $T_{min}$ + I($T_{min}$^2) + DBH + I(DBH^2)",  random = ~1|coreID, correlation = corCAR1(form=~yeaar|coreID), data = x, na.action = "na.fail", method = "ML")*

where *x* is a complete data set (with no missing values) for one species at one site. The method is set to maximum likelihood (*ML*) during the model selection phase, but to restricted maximizing likelihood (*REML*) once the best model is identified.

@@

To test for yeaar effects, we limited the analysis to species with reasonable coverage of the DBH x `year` matrix.
Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total DBH range for $\ge$ 2/3 of the total time range analyzed. 
To avoid severe big-tree selection biases [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm, if present.
Species that failed to meet these criteria (n= 8) were excluded from the analysis of temporal trends.
We then ran models as described above, including a  first-order linear effect of year.
We note that the random effect of tree should in theory avoid analytical biases arising from persistent growth differences among individuals [@brienen_detecting_2012; @brienen_tree_2017]. 
To verify that GLS model trends for yeaar were real, as opposed to an artifact of inherent covariation between DBH and yeaar within each core, we compared GLS results to an analysis of DBH-growth relationships by decade (Figs. `r decadal_analysis_figs_all`). 
For simplicity and consistency with previous literature, we present only the models with BAI as the growth metric, noting that responses were similar across growth metrics.

Within each of three categories of models run (climate only, climate + DBH, climate $\times$ DBH, climate + DBH + year), we selected as the top model that with the lowest AIC.

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration (PET). Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R, @zang_dendroclimatic_2013]. Black rectangle represents the period selected by *climwin* as the most influential window. Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta$AIC for each. The time window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the optimal window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model). Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Species names corresponding to the codes are given in Table `r species_table`.](tables_figures/quilt_comparison.png)

@@

```{r , eval=FALSE, echo=FALSE}
## Tallies for the following paragraph (Feb 11)
#sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
#sameV_diffW: BCNMp, BCNMt, HFt, HKKp, LDWt,  SCBIt, SCp,SCt,  CBp, CBt (10)
#diffV_diffW: HFp, HKKt
```
The three metrics of growth (RW, BAI, and $\Delta$AGB)  exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs. `r all_climwin`).
In 8 of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_SCBI_W`). 
In 10 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_SCBI_T`). 
Finally, in two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF), *climwin* identified different climate variables and different time windows (Figs. `r climwin_HKK_T`, `r climwin_HF_T`). 
Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with RW as the growth metric and for the full set of cores (*i.e.*, including those for which DBH could not be reconstructed).

@@

![**Figure 3 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** *Climwin*-selected climate variables are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential. For each species (color-coded as in Fig. 4), relationships are plotted if included in top model. For each relationship shown, other terms in the model are held constant at their median. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)

@@

Interactive effects of climate and DBH were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-variable combinations for which they were tested. 
For precipitation variables, interactions were significant for `r n_DBHxP_RW_sig` of the `r n_DBHxP_RW_analyzed` (`r round(n_DBHxP_RW_sig/ n_DBHxP_RW_analyzed*100)`%) interactions with RW as the growth metric (Fig. `r RW_interactions_all`) and for `r n_DBHxP_BAI_sig` of the `r n_DBHxP_BAI_analyzed` (`r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`%) with BAI as the growth metric (Fig. `r BAI_interactions_all`).
The majority of these interactions were positive (`r round(n_DBHxP_RW_sig_pos/n_DBHxP_RW_sig*100)`% for RW; `r round(n_DBHxP_BAI_sig_pos/n_DBHxP_BAI_sig*100)`% for BAI), indicating that larger trees generally respond more positively to precipitation or its frequency (Fig.  `r interactions_fig`).
Among the exceptions to this pattern (n=`r n_DBHxP_RW_sig-n_DBHxP_RW_sig_pos` for RW, `r n_DBHxP_BAI_sig-n_DBHxP_BAI_sig_pos` for BAI), only a minority (n=1 for RW, 4 for BAI) occurred in species responding positively to precipitation in the current growing season.

Temperature variable $\times DBH$ interactions were significant for `r round(n_DBHxT_RW_sig/ n_DBHxT_RW_analyzed*100)`% of cases with RW as the growth metric (Fig. `r RW_interactions_all`) and for `r round(n_DBHxT_BAI_sig/ n_DBHxT_BAI_analyzed*100)`% with BAI as the growth metric (Fig. `r BAI_interactions_all`).
Directions of these interactions were mixed, with `r n_DBHxT_RW_sig_neg` of `r n_DBHxT_RW_sig` significant interactions negative with RW as the growth metric and `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative when BAI was the growth metric.
For both RW and BAI, the majority of significant negative interactions (*i.e.*, more negative/ less positive response of larger trees to higher temperatures) occurred in cases where the main effect temperature was negative (e.g., HKK, LT, CB; Fig.  `r interactions_fig`), whereas positive interactions were more common when the main effect of temperature was positive (e.g., HF, ZOF).

@@

When a precipitation variable, a temperature variable, and DBH were all included as candidate variables in the GLS models, typically all three were included in the top model, regardless of the growth metric used.
Climate responses were generally similar to those described above for models without a DBH term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCNM; Figs. `r clim_sens`, `r all_GLS`).
In general, DBH explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r all_GLS`).
The relative importance of DBH tended to be least for RW, intermediate for BAI, and highest for $\Delta$AGB (e.g., at SCBI; Fig. `r GLS_SCBI`).
However, there were exceptions, particularly when RW decreased steeply with DBH (e.g., LT; Fig. `r GLS_LT`).

@@

All three growth metrics, RW, BAI, and $\Delta$AGB, varied with DBH for most species at all sites (Fig. `r size`).
For RW, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear DBH term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative). 
There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
On one end of the spectrum, some species exhibited maximum RW at low DBH, followed by fairly rapid declines in RW with increasing DBH. 
Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) and/or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig. `r size`).
At the other end of the spectrum, some species had low RW at small DBH, increased to peak RW at intermediate DBH, and subsequently declined.
These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp. at SCBI and Žofín, *Acer saccharum* at LDW, *Picea* spp. at Žofín and CB; Table `r species_table`).
